---
title: "ANSWERS: Working with untidy and tidy data"
subtitle: "Data Science for Biologists, Spring 2020"
author: "Madeline Greene"
output: 
  html_document:
    highlight: tango
css: static/sjs-style.css
params:
  data_path: "data/"
---

<!-- SETUP CODE CHUNK -->
```{r setup, include=FALSE}
## Change the default figure width and height to your liking, but within reason.
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=3.5)
library(tidyverse)
library(kableExtra) # helpful to print fancy tables in RMarkdown
theme_set(theme_classic())  ### If you want to set a global theme, do so here
```


# <mark>Questions</mark>

#### Preface Question

Look at the YAML front matter: there is something NEW! Read about params [here](https://rmarkdown.rstudio.com/lesson-6.html). In this case we set a global parameter for the RMarkdown file that corresponds to the path where data lives! We can proceed to use this as a variable like so (always `params$`, like a column in a data frame!). Use this variable for your paths, e.g. `read_csv(paste0(params$data_path, <NAMEOFFILE>))`
```{r}
params$data_path

```

#### Question 1

```{r}
# Code chunk to read in and examine the data with head()
read_csv(paste0(params$data_path, "mcv_immunized_percent_of_one_year_olds.csv")) -> mcv_immunized #other way is to rename the name of the file 
head(mcv_immunized) #just viewing head
```


```{r}
# Code chunk to convert data to tidy
mcv_immunized %>%
  pivot_longer(`1980`:`2011`, names_to = "year", values_to = "percent") -> tidymcv

tidymcv


```
#### Question 2


```{r, fig.width = 11, fig.height = 6}
# Code chunk to plot the data
tidymcv %>%
  filter(country %in% c("Brazil","Central African Republic")) %>%
  #comparing these two countries
  ggplot(aes(x =year, y = percent, group = country)) +
  geom_line() + 
  geom_point() +   
  facet_wrap(~country, nrow=2)+
  #nrow = overtop of eachother
  theme(strip.background = element_rect(fill = "coral")) +
  labs(x = "Year", y = "Percent")
  
```

<answer>
Based on this line graph, it is evident that Brazil has an overall higher percent of vaccinations. Central African Republic has a few years where it climbed in percentage from 1987-1990, then dropped down at 1992, and remained pretty steady at around 50% until 2011. Brazil is mainly over 75% vaccinated, almost reaching a full 100%.
</answer>

#### Question 3


```{r}
# Code chunk to read in and examine the data with head()
read_csv(paste0(params$data_path, "dtp3_immunized_percent_of_one_year_olds.csv")) -> dtp3_immunized
dtp3_immunized
head(dtp3_immunized)

```


```{r}
# Code chunk to convert data to tidy and combine with question 1 data
dtp3_immunized %>%
  pivot_longer(`1980`:`2011`, names_to = "year", values_to = "percent") -> tidy_dtp3
  #need back ticks to make it work
tidy_dtp3 %>%
  mutate(vaccine = "dtp3") -> dtp3_vaccine

tidymcv %>%
  mutate(vaccine = "mcv") -> mcv_vaccine

bind_rows(mcv_vaccine, dtp3_vaccine) -> new_vaccine
#have the same labels so can bind
new_vaccine

new_vaccine %>%
  mutate(year = as.numeric(year)) -> numeric_vaccine
  #want year
numeric_vaccine



```

```{r}
# Code chunk to plot the data
numeric_vaccine %>%
  filter(country == "Brazil") %>%
  #selected Brazil
  ggplot(aes(x = year, y= percent, fill = vaccine)) +
  geom_col(position = "dodge") + 
  #wanted the columns next to eachother
  labs(x = "Year", y = "Percent", fill = "Vaccine") +
  scale_fill_brewer(palette= "Dark2")

```

<answer>
It is evident that in this figure, it shows that in the country Brazil, from 1980 to 2010 roughly, the percent of vaccinations definitely increased. It increased significantly around 1990.
</answer>

#### Question 4


```{r}
# Code chunk to read in and examine the data with head()
read_csv(paste0(params$data_path, "seals.csv")) -> seals
seals
head(seals)
```


```{r}
# Code chunk for distributions (violin, sina, etc.)
# If sina make sure to load `ggforce()` in SETUP CHUNK!

seals %>%
  ggplot(aes(x = feeding, y = oxygen_use)) +
  geom_violin() +
  #violin plot shows distribution
  labs(x = "Feeding", y = "O2 Usage ")
 
  
```

```{r}
# Code chunk for data un-tidying and scatterplot

seals %>%
  pivot_wider(names_from = "feeding", values_from = "oxygen_use") -> seals_feeding

seals_feeding %>%
  ggplot(aes(x = No, y = Yes)) +
  geom_point() +
  geom_abline(col = "pink") +
  #added y=x line
  labs( x = "Oxygen Use When NOT Feeding", y = "Oxygen Use When Feeding") +
  ylim(25,150) +
  #made y limit
  xlim(25,150) +
  #made x limit
  annotate("text", x = 60, y = 50, label = "y=x")
  #added text annotation
  

```

<answer>
Based on both figures, the oxygen use is increased WHEN feeding. You can tell this by the scatterplots positive increase. You can also tell this by the violin plot since the distribution is higher for oxygen use when feeding.
</answer>

#### Question 5


```{r}
# Code chunk to read in and examine the data with head()
read_csv2(paste0(params$data_path, "planets.csv")) -> planets
planets
head(planets)

```


```{r}
# Code chunk for tidying
# HINT: it's literally just piping data into separate_rows twice, one for the column given in the question, and for the second you need to figure out the column.
## DO NOT OVERCOMPLICATE THIS QUESTION.

planets %>%
  separate_rows(terrain) %>%
  #separated terrain because had multiple names
  separate_rows(climate) ->tidy_planets
  #separated climate because had multiple names
head(tidy_planets)
```


#### Question 6

```{r}
# Code chunk for allllll the wrangling and tidying
# should END UP as a single piped command - but build it up LINE AT A TIME!

tidy_planets %>%
  separate(edited, c("edited_date", "edited_time"), sep = "T") %>%
  #need to separate with "T" because that is where it separates the date and time
  mutate(ratio_period_rotation_orbital = rotation_period/orbital_period) %>%
  #dividing is way for ratio
  unite("name_pop", name, population, sep = "-") %>%
  #sep with a dash
  filter(terrain == "mountains") %>%
  #only want terrain
  arrange(desc(ratio_period_rotation_orbital)) -> mountain_planets
  #want it descending

mountain_planets
  
```


#### Question 7


```{r}
## defines each column as an array
tibble(
  color        = c("blue", "green", "turquoise"),
  male         = c(25, 36, 52),
  female       = c(28, 17, 33),
  species_abbr = c("sp1", "sp1", "sp1")
) -> frogs_sp1

## defines tibble in a format like the tibble would end up appearing
## spacing is helpful!
tribble(
  ~color, ~male, ~female, ~species_abbr,
  "orange", 13,   17,    "sp2",
  "red",    45,   36,    "sp2",
  "pink",  31,    25,    "sp2"
) -> frogs_sp2


## Like strategy one
tibble(
  species_name = c("Frogus firstus", "Frofus secondus"),
  species_abbr = c("sp1", "sp2")
) -> species_names

### ADD CODE HERE TO PRINT OUT THESE THREE DATASETS!!! YOU HAVE TO LOOK AT THEM TO USE THEM!!

frogs_sp1

frogs_sp2

species_names
```



```{r}
# Code chunk to tidy and combine the three datasets

frogs_sp1 %>%
  bind_rows(frogs_sp2)%>%
  #the data is more or less the same so binded the rows
  left_join(species_names) -> tidy_frog
  #tried right and left join and didn't see much of a difference so just left join

tidy_frog #printed to make sure looked okay

tidy_frog %>%
  pivot_longer(male:female, names_to = "Frog_sex", values_to = "Frog_count") %>%
  select(-species_abbr) %>%
  #told to remove this column
  separate(species_name, c("Genus", "Species")) -> frog_tidy_tibble

frog_tidy_tibble

```

```{r}
# Code chunk to for dplyr task 1
frog_tidy_tibble %>%
  filter(Species == "firstus") %>%
  #filtered by species : Frogus firstus
  filter(Frog_sex == "female") %>%
  filter(color %in% c("blue", "green")) 

#blue have more

```

```{r}
# Code chunk to for dplyr task 2
frog_tidy_tibble %>%
  filter(Frog_sex == "male") %>%
  #only wanted male
  group_by(Species) %>%
  #wanted to show species
  summarize(sum_males = sum(Frog_count))
  #wanted overall sum between species

```

```{r}
# Code chunk to for dplyr task 3
frog_tidy_tibble %>%
  filter(Species == "secondus") %>%
  select( -Genus) %>%
  select(- Species) %>%
  pivot_wider(names_from = "Frog_sex", values_from = "Frog_count") %>%
  mutate(ratio_frogs = male/female) %>%
  select(color, ratio_frogs) %>% 
  arrange(desc(ratio_frogs))

#Red Frogus secondus have the highest male:female ratio!!


```



#### Question 8

```{r}
# Code chunk for kable()

frog_tidy_tibble %>%
  kable() %>%
  kable_styling(bootstrap_options = "hover", full_width = F, font_size = 10, position = "left") %>%
  #wanted to be able to hover over each line, towards left of page
  row_spec(1:2, color = "blue") %>%
  row_spec(3:4, color = "green") %>%
  row_spec(5:6, color = "turquoise") %>%
  row_spec(7:8, color = "orange") %>%
  row_spec(9:10, color = "red") %>%
  row_spec(11:12, color = "pink") -> kable_frog
  #made each color word the color it is

kable_frog
```

