---
title: "Introduction to linear regression"
subtitle: "Data Science for Biologists, Spring 2020"
author: "Madeline Greene"
output: 
  html_document:
    highlight: tango
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(broom) 
```

## Instructions

Standard grading criteria apply, except there is no "answer style" - just write out answers normally! Note the following:

+ When adding regression lines as plot subtitles, don't worry about writing $\epsilon$
+ Do not remove insignificant predictors from your regression line formula when reporting it (this is just a question I get a lot)

This assignment will use an external dataset from a field experiment studying the diversity of [Chinese Rowan](https://en.wikipedia.org/wiki/Rowan) trees. Researchers randomly sampled and recorded characteristics of leaves from three different species in the *Sorbus* genus. They recorded the following measurements for each tree (don't worry about units)

1. `species`: the species of tree
2. `altitude`: the altitude of the tree
3. `respiratory_rate`: average respiratory rate across a random sample of leaves from that tree
4. `leaf_len`: average leaf length across a random sample of leaves from that tree
5. `birds_nesting`: whether birds were actively nesting in the tree

For this assignment, you will examine how various predictors may explain variation in *respiratory rate.*

```{r}
# rowan trees, no relation :)
rowan <- read_csv("https://raw.githubusercontent.com/sjspielman/datascience_for_biologists/master/data/rowan_trees.csv")
dplyr::glimpse(rowan)
```

### Question 1


> Background for this completing question is under the header "Simple linear regression: Single numeric predictor" in the linear regression tutorial


Construct a linear model that examines how *altitude* might explain variation in *respiratory rate*. Take the following steps to do so (code as you go through the steps!)

Make a quick scatterplot to make sure the "linear relationship" assumption is met for this data. Be sure to have your predictor and response on the correct axis!:

```{r}
### figure to check linearity goes here.
### no need for fancy, can keep labels as is, etc.

ggplot(rowan, aes(x = altitude, y = respiratory_rate)) + 
  geom_point()
```


Now that you have confirmed the relationship is linear (hint: it is linear), build your linear model. *BEFORE you examine its output*, evaluate whether the model residuals were normally distributed:

```{r}
### build model and check normal residuals
### do not look at the model summary yet!

lm(respiratory_rate ~ altitude, data = rowan) -> fit
# Y~X

#get the residuals
broom::augment(fit, rowan) -> augmented_fit
qqnorm(augmented_fit$.resid, pch = 19)
qqline(augmented_fit$.resid, col = "pink")
#pch = 19 is closed circles, pch = 21 is open circles
```


Now that you have confirmed the residuals are roughly normally distributed (hint: they are), examine the output from your linear model. In the space below the code chunk, discuss in *bullet form* (1-2 sentences each): a) Provide an interpretation of the intercept, b) Provide an interpretation of the altitude coefficient, c) Provide an interpretation of the $R^2$ value (those dollar signs signify "math mode" - see the HTML output!), and finally d) Conclude whether altitude is a strong or weak predictor of respiratory rate, consider "biological significance" (effect size!) as well as statistical significance.

```{r}
## examine model output here
summary(fit)
```

+ Intercept interpretation : We expect that a tree's respiratory rate will be 8.386e-03 when it's altitude is 0. However, p it not significant at P = 0.622 so we actually have no evidence that the intercept differs from 0.

+ altitude coefficient interpretation : For every unit increase in altitude, we expect the respiratory rate to increase by 3.025e-04. It is highly significant at P =2.66e-14, so we reject the null that the coefficient is 0.

+ $R^2$ interpretation : $R^2 = 0.1743$ means ~ 17% of variation in the respiratory rate is explained by our model, which included only the altitude predictor. It is highly significant at 2.66e-14. This means that ~83% of variation in the respiratory rate is unexplained. 

+ Model conclusion : It is a significant model but has a weak explanatory power. 




Finally, make a stylish scatterplot of your findings. Your scatterplot should:

+ Use your favorite ggplot theme and colors (it's allowed to like the default!)
+ Clearly show the regression line and its 95% confidence interval
+ Include a meaningful title, along with a subtitle that is the fitted model itself, as well as other nice labels
+ Include a well-placed annotation that gives the model's $R^2$

```{r}
### stylish plot goes here
formula = "Y = 3.025e-04X + 8.386e-03"
ggplot(rowan, aes(x = altitude, y = respiratory_rate)) +
  #provided individual data points
  geom_point() +
  #title of graph and equation
  labs(title = "Regression of Altitude Explaining Respiratory Rate",
       subtitle = "Y = 3.025e-04X + 8.386e-03") +
  labs(x = "Altitude", y = "Respiration Rate") +
  #used "lm" to have strait line
  #level = 95% confidence 
  geom_smooth(method = "lm", col = "pink") +
  #provided R2 value in white space of figure
  annotate("text", 
           x = 200, y = 0.3, 
           label = "R^2 == 0.1743", 
           parse = TRUE)
```



### Question 2

> Background for this completing question is under the header "Simple ANOVA: Single categorical predictor" in the linear regression tutorial

Construct a linear model that examines how *species* might explain variation in *respiratory rate*. Take the following steps to do so (code as you go through the steps!)

Make a quick plot (use `geom_point()`, seriously, not even a jitter!!) to make sure the "equal variance" assumption is met for this data:

```{r}
### figure to check assumption goes here.
### no need for fancy, can keep labels as is, etc.

ggplot(rowan, aes(x = species, y = respiratory_rate)) +
  geom_point()
```


Now that you have confirmed the variance is equal across groups (hint: it is), build your linear model. *BEFORE you examine its output*, evaluate whether the model residuals were normally distributed:

```{r}
### build model and check normal residuals
### do not look at the model summary yet!

lm(respiratory_rate ~ species, data = rowan) -> fit_species
#get the residuals
broom::augment(fit_species, rowan) -> augmented_fit
qqnorm(augmented_fit$.resid, pch = 19)
qqline(augmented_fit$.resid, col = "pink")
#pch = 19 is closed circles, pch = 21 is open circles
```

Now that you have confirmed the residuals are roughly normally distributed (hint: they are), examine the output from your linear model. In the space below the code chunk, discuss in *bullet form* (1-2 sentences each): a) Provide an interpretation of the intercept, b) Provide an interpretation of the species coefficient, c) Provide an interpretation of the $R^2$ value, and finally d) Conclude whether species is a strong or weak predictor of respiratory rate, consider "biological significance" (effect size!) as well as statistical significance.

```{r}
## examine model output here
summary(fit_species)
```

+ Intercept interpretation : for microphylla trees, we expect a mean resp rate of 0.148680. highly significant at  p<2e-16, intercept definitely not 0. 

+ species coefficient interpretation :
  + We expect on average, oligodonta trees to have a respiration rate 0.018830 lower than microphylla. However, it is not significant, therefore there is no evidence it differs from 0.
  + We expect on average, sargentiana trees to have a respiration rate 0.006180 lower than microphylla. However, it is not significant, therefore there is no evidence it differs from 0.
  
+ $R^2$ interpretation : $R^2 = 0.002792$ which is not significant at P = 0.2437, concludes that there is no evidence that any variation is explained. 100% of the variation in the respiration rate is unexplained by our model.

+ Model conclusion : It is entirely an uninformative model - species does NOT predict/explain respiration rate according to model.

Finally, make a stylish figure of your findings, choosing your own geom!

+ Use your favorite ggplot theme and colors (it's allowed to like the default!)
+ If your geom does not already show the center of each group (i.e. like a boxplot), be sure to add the means in with `stat_summary()`
+ Include a meaningful title, along with a subtitle that is the fitted model itself, as well as other nice labels
+ Include a well-placed annotation that gives the model's $R^2$

```{r}
### stylish plot goes here

ggplot(rowan, aes(x = species, y = respiratory_rate)) + 
  geom_jitter(width = 0.2) +
  #puts the mean number as a point, colored it
  stat_summary(color = "coral") +
  labs( x = "Species", 
        y = "Respiration Rate") +
  labs(title = "Altitude Explaining Respiration Rate Per Species" , 
       subtitle = "Y = -0.018830X1 + -0.006180X2 + 0.148680") + 
  #put R2 in whitespace 
  annotate("text", 
           x = 1.5, y = 0.3, 
           label = "R^2 == 0.002792", 
           parse = TRUE)
```





### Question 3


> Background for this completing question is under the header "LM with numeric and categorical predictors" in the linear regression tutorial


Construct a linear model that examines how BOTH *species* and *leaf_len* as independent effects might explain variation in *respiratory rate*. Again, take the following steps one by one:

Make TWO quick plots to make sure assumptions are met for each of our predictors:

```{r}
### figures to check assumption goes here.
### no need for fancy, can keep labels as is, etc.

ggplot(rowan, aes(x = leaf_len, y = respiratory_rate)) + geom_point()
```


Build your linear model, and evaluate whether the model residuals were normally distributed:

```{r}
### build model and check normal residuals
### do not look at the model summary yet!

lm(respiratory_rate ~ species + leaf_len, data = rowan) -> fit_leaf
#get the residuals
broom::augment(fit_leaf, rowan) -> augmented_fit
qqnorm(augmented_fit$.resid, pch = 19)
qqline(augmented_fit$.resid, col = "pink", lwd = 5)
#pch = 19 is closed circles, pch = 21 is open circles
```

Now that you have confirmed the residuals are roughly normally distributed (hint: they are), examine the output from your linear model. In the space below the code chunk, discuss in *bullet form* (1-2 sentences each): a) Provide an interpretation of the intercept, b) Provide an interpretation of the `species` coefficient, c) Provide an interpretation of the `leaf_len` coefficient d) Provide an interpretation of the $R^2$ value, and finally e) Conclude whether species is a strong or weak predictor of respiratory rate, consider "biological significance" (effect size!) as well as statistical significance.

```{r}
## examine model output here

summary(fit_leaf)
```

+ Intercept interpretation : For a microphylla tree that has a leaf length of 0, we expect for it's respiration rate to be 0.176, which is highly significant. While it is significant, it is not biologically realistic information.

+ species coefficient interpretation :
  + When controlling for leaf length, the coefficients are the change from the baseline species.
  + When controlling for lead length, the oligodonta species have an average respiration rate of 0.003839 lower than microphylla, but not significant so effectively 0. This means that there is no difference from microphylla.
+ leaf_len coefficient interpretation : When controlling for species,it is expected that for every additional unit of the leaf length,  the respiration rate will decrease 0.003054X3, demonstrating a negative slope.
+ $R^2$ interpretation : $R^2 = 0.01398$ is significant, which means 100% of the respiration rate can be explained by the leaf length.
+ Model conclusion : The model conclusion is that the model has no explanatory power.

Finally, make a stylish scatterplot of your findings:

+ Use your favorite ggplot theme and colors (it's allowed to like the default!)
+ Make sure to show a regression lines for EACH species. **NOTICE in theses lines** how they are consistent with your conclusions about species being a significant predictor. You do not have to write anything, just notice!
+ Include a meaningful title, along with a subtitle that is the fitted model itself, as well as other nice labels
+ Include a well-placed annotation that gives the model's $R^2$

```{r}
### stylish plot goes here

ggplot(rowan, aes(x = leaf_len, y = respiratory_rate, color = species))+
  geom_point() +
  #created regression line
  geom_smooth(method = "lm") +
  #I realize not colorblind friendly but wanted to try it
  scale_color_brewer(palette = "Accent") +
  labs( x = "Leaf Length", 
        y = "Respiration Rate", 
        color = "Species")+
  #had title plus equation
  labs(title = "Regression of Leaf Length Explaining Respiratory Rate Per Species", 
       subtitle = "Y = -0.003898X1 +  0.050740X2 + -0.003054X3 + 0.175923") +
  #put the annotation in the white space
  annotate("text", 
           x = 35, y = 0.35, 
           label = "R^2 == 0.01398", 
           parse = TRUE)

```

